---
title: "TB Incidence Around the Globe"
author: "Di Cook"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: "css/logo.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  error = FALSE)
```


## Data description

This is current tuberculosis data taken from [WHO](http://www.who.int/tb/country/data/download/en/), the case notifications table. There's also an R package, [TBinR](https://cran.r-project.org/web/packages/getTBinR/index.html), that can be used to dynamically download this data.

```{r read TB data and wrangle and subset to USA}
library(tidyverse)
tb <- read_csv(here::here("data/TB_notifications_2019-07-01.csv")) %>% 
  select(country, iso3, year, g_whoregion, new_sp_m04:new_sp_fu) %>%
  gather(stuff, count, new_sp_m04:new_sp_fu) %>%
  separate(stuff, c("stuff1", "stuff2", "sexage")) %>%
  select(-stuff1, -stuff2) %>%
  mutate(sex=substr(sexage, 1, 1), 
         age=substr(sexage, 2, length(sexage))) %>%
  select(-sexage) %>%
  filter(year > 1996, year < 2013) %>%
  filter(!(age %in% c("04", "014", "514", "u")))
ggplot(tb, aes(x=g_whoregion)) +
  geom_bar(aes(weight = count))
```

### Quiz

```{r quiz}
quiz(
  question("What region of the world has the highest incidence?",
    answer("southeast Asia", correct = TRUE),
    answer("Africa"),
    answer("Europe"),
    answer("America"),
    answer("Western Pacific"), 
    answer("Eastern Mediterranean"))
)
```


## Exercises 

Write the R code required to filter the platypus data to contain only incidence for 2010.

```{r tb2010, exercise=TRUE}
glimpse(tb)
```

```{r tb2010-solution}
tb %>% filter(year >= 2010)
```


## Temporal trends

This is a an example of an interactive chart that allows exploring the temporal trend of TB incidence in India, with a scrollbar to cycle through different age groups.

```{r fig.width=6, fig.height=6}
library(plotly)
p <- tb %>% filter(iso3 == "IND") %>% 
  filter(!is.na(count)) %>%
  ggplot() + 
  geom_point(aes(x=year, y=count, colour=sex, frame=age, label=iso3)) +
  geom_smooth(aes(x=year, y=count, colour=sex, frame=age),  se=FALSE) + 
  scale_colour_brewer(palette = "Dark2") +
  ylim(c(0, 100000))
ggplotly(p)
```
