---
title: "TB Incidence Around the Globe"
author: "Di Cook"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: "css/logo.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  error = FALSE)
```


## Data description

This is current tuberculosis data taken from [WHO](http://www.who.int/tb/country/data/download/en/), the case notifications table. There's also an R package, [TBinR](https://cran.r-project.org/web/packages/getTBinR/index.html), that can be used to dynamically download this data.

```{r read TB data and wrangle and subset to USA}
library(tidyverse)
library(DT)
tb <- read_csv(here::here("data/TB_notifications_2019-07-01.csv")) %>% 
  dplyr::select(country, iso3, year, new_sp_m04:new_sp_fu) %>%
  pivot_longer(cols=new_sp_m04:new_sp_fu, names_to="sexage", values_to="count") %>%
  mutate(sexage = str_replace(sexage, "new_sp_", "")) %>%
  mutate(sex=substr(sexage, 1, 1), 
         age=substr(sexage, 2, length(sexage))) %>%
  dplyr::select(-sexage)  %>%
  filter(country == "United States of America") %>%
  filter(!(age %in% c("04", "014", "514", "u"))) %>%
  filter(year > 1996, year < 2013) 

datatable(tb)
```

## Incidence

```{r plot incidence by region}
ggplot(tb, aes(x=year)) +
  geom_bar(aes(weight = count)) 
```

```{r quiz}
quiz(
  question("How would you describe the trend in TB incidence?",
    answer("decreasing", correct = TRUE),
    answer("increasing"),
    answer("flat"),
    answer("volatile"))
)
```


## Exercises 

Write the R code required to filter the TB data to contain only incidence for 2010.

```{r tb2010, exercise=TRUE}
glimpse(tb)
```

```{r tb2010-solution}
tb %>% filter(year >= 2010)
```


## Temporal trends

This is a an example of an interactive chart that allows exploring the temporal trend of TB incidence in India, with a scrollbar to cycle through different age groups.

```{r fig.width=6, fig.height=6}
library(plotly)
p <- tb %>% 
  group_by(year, age) %>%
  summarise(p_males = count[sex=="m"]/sum(count)) %>%
  ggplot() +
    geom_point(aes(x=age, y=p_males, frame=year))
ggplotly(p)
```
