---
title: "Where do I find a platypus in Australia"
subtitle: "outside of the zoo"
author: "Di Cook"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: "css/logo.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  error = FALSE)
```


## Data description

The platypus is a very strange animal, one of just two mammals in the worl to lay eggs! They are found exclusively in Australia. 

The data used in this tutorial is downloaded from [Atlas of Living Australia](https://www.ala.org.au), downloading occurrences of "Ornithorhynchus anatinus":

```{r}
library(tidyverse)
library(ggmap)
load(here::here("data/platypus.rda"))
load(here::here("data/oz.rda"))
platypus <- platypus$data
ggmap(oz) + 
  geom_point(data=platypus, 
             aes(x=longitude, y=latitude), 
              alpha=0.1, colour="orange") 
```

### Quiz

```{r quiz}
quiz(
  question("What part of Australia is the platypus natural habitat?",
    answer("southeast", correct = TRUE),
    answer("west"),
    answer("north"),
    answer("centre")
  )
)
```

## Exercises 

Write the R code required to filter the platypus data to contain sightings since 2010.

```{r since-2010, exercise=TRUE}
glimpse(platypus)
```

```{r since-2010-solution}
platypus %>% filter(year >= 2010)
```

## Interactive plot

```{r fig.width=6, fig.height=6}
library(plotly)
library(ggthemes)
platy_2010 <- platypus %>% filter(year >= 2010)
p <- ggmap(oz) + 
  geom_point(data=platy_2010, 
             aes(x=longitude, y=latitude), 
              alpha=0.1) +
  geom_point(data=platy_2010, 
             aes(x=longitude, y=latitude, frame = year), 
              alpha=0.5, colour="orange") +
  theme_map()
ggplotly(p)
```



